---
title: "Empty blogpost template"
author: "Ivann Schlosser"
 html_document: 
    code_folding: show
    toc: yes
    toc_float: yes
    css: style_rmd.css
    number_sections: yes
    code_download: yes
    highlight: pygments
    fig_width: 6
    fig_height: 4
    fig_caption: yes
    df_print: kable
    theme: yeti
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, echo=FALSE}

library(formatR)
knitr::opts_chunk$set(echo = TRUE
                      ,eval = FALSE
                      ,include = TRUE
                      ,message = FALSE
                      ,warning = FALSE
                      ,fig_caption = FALSE
                      ,out.width = "80%"
                      ,fig.align = "center"
                      # ,tidy = TRUE
                      ,cache = FALSE)

```

# Intro

This blog post will cover exception handling in R. The main purpose of it is to ultimately build robust, fault proof scripts, that can be part of your everyday workflow, but also be easily shared with others, via packages or version control and perform as expected in any situation or raise informative messages to the user as to the possible problems.

# Context

Let's look at the output of the following operation, it is intentionally misused to raise an error message:

One of the most important elements of any good software is it's ability to put the user in a situation, where it is almost impossible to produce an error, and if an error still occurs somewhere, then the message associated to it should allow the user to straight away correct it.

# Examples

```{r}

addition <- functino(x,y) {
  return(x+y)
}

addition(1, "number")

```


# Sources
This post is a compilation of personal experience and reading from these amazing sources:

* [Advanced R](https://adv-r.hadley.nz/debugging.html)
* [GeeksForGeeks](https://www.geeksforgeeks.org/handling-errors-in-r-programming/)
* [Mastering software development in R](https://bookdown.org/rdpeng/RProgDA/error-handling-and-generation.html)


